The Parareal Algorithm for Parallel Solution of ODEs

### 2014-12-09 paper.tex
changed poster and writeup

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We implement the parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
differnt solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for variuos complexities of ODEs from the basic exponential to
modelling sound wave propagation as we did in Homework 4.
\end{abstract}
%\end{@twocolumnfalse}
%]

>>>>

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We implement the parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
different solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for various complexities of ODEs from the basic exponential to
modeling sound wave propagation as we did in Homework 4.
\end{abstract}
%\end{@twocolumnfalse}
%]

### 2014-12-09 poster/main.tex
changed poster and writeup

\usepackage{graphicx}  % Required for including images

\usepackage{booktabs} % Top and bottom rules for tables
\usepackage[linesnumbered,vlined,ruled]{algorithm2e} % algorithm

@@

We implement the parareal algorithm to solve various ODEs and compare in terms
of accuracy, speedup and efficiency. We explore the effect of using differnt
solution operators (a coarse and a fine as required by the method) The majority
of our code is written in python using mpi4py to parallelize.  To compare to
more performance-based languages, we compare certain codes to an implementation
in C++.

@@

\begin{block}{Introduction}

Solving systems of differential equations can be a computationally expensive
task. The error of most algorithms scales on the stepsize of the discretization
of time. However, stepsize in time is also proportional to the computation
required. The parareal algorithm, developed by YADA YADA. An alternative would
be to parallelize by space.

%------------------------------------------------

\begin{figure}
\includegraphics[width=0.8\linewidth]{harvard_univ_logo.png}
\caption{Figure caption}
\end{figure}

@@

\begin{column}{\twocolwid} % Begin a column which is two columns wide (column 2)

\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column

\begin{column}{\onecolwid}\vspace{-.6in} % The first column within column 2 (column 2.1)

@@

\begin{block}{Algorithm}
The general parareal method, as described by YADA YADA can be SEEN BELOW YAR

\begin{enumerate}
    \item Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$
    \item Compute the corrections $\delta g_n(u_1_n) =
        g_{\textnormal{fine}}(t_n, u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel
    \item Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta
        t}(t_n, u^2_n) + \delta g_n(u^1_n)$
    \item Repeat steps 2 and 3, incrementing the iteration label and using
        $u^{k+1}_0 = u^1_0$ as the initial condition
\end{enumerate}
\end{block}

@@

\begin{column}{\onecolwid}\vspace{-.6in} % The second column within column 2 (column 2.2)

Lorem ipsum dolor \textbf{sit amet}, consectetur adipiscing elit. Sed laoreet accumsan mattis. Integer sapien tellus, auctor ac blandit eget, sollicitudin vitae lorem. Praesent dictum tempor pulvinar. Suspendisse potenti. Sed tincidunt varius ipsum, et porta nulla suscipit et. Etiam congue bibendum felis, ac dictum augue cursus a. \textbf{Donec} magna eros, iaculis sit amet placerat quis, laoreet id est. In ut orci purus, interdum ornare nibh. Pellentesque pulvinar, nibh ac malesuada accumsan, urna nunc convallis tortor, ac vehicula nulla tellus eget nulla. Nullam lectus tortor, \textit{consequat tempor hendrerit} quis, vestibulum in diam. Maecenas sed diam augue.

@@

\end{columns} % End of the split of column 2 - any content after this will now take up 2 columns width

@@

\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column again

\begin{column}{\onecolwid} % The first column within column 2 (column 2.1)

@@

%\begin{block}{Mathematical Section}

%Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.

%\begin{equation}
%E = mc^{2}
%\label{eqn:Einstein}
%\end{equation}
%
%Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.
%
%\begin{equation}
%\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
%\label{eq:refname}
%\end{equation}
%
%Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.
%
%\begin{equation}
%\kappa =\frac{\xi}{E_{\mathrm{max}}} %\mathbb{ZNR}
%\end{equation}
%
%\end{block}

@@

@@

\begin{block}{Results}

\begin{figure}
\includegraphics[width=0.8\linewidth]{harvard_univ_logo.png}
\caption{Figure caption}
\end{figure}

Nunc tempus venenatis facilisis. Curabitur suscipit consequat eros non porttitor. Sed a massa dolor, id ornare enim:

\begin{table}
\vspace{2ex}
\begin{tabular}{l l l}
\toprule
\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
\midrule
Treatment 1 & 0.0003262 & 0.562 \\
Treatment 2 & 0.0015681 & 0.910 \\
Treatment 3 & 0.0009271 & 0.296 \\
\bottomrule
\end{tabular}
\caption{Table caption}
\end{table}

@@

\end{columns} % End of the split of column 2

\end{column} % End of the second column

@@

Nunc tempus venenatis facilisis. \textbf{Curabitur suscipit} consequat eros non porttitor. Sed a massa dolor, id ornare enim. Fusce quis massa dictum tortor \textbf{tincidunt mattis}. Donec quam est, lobortis quis pretium at, laoreet scelerisque lacus. Nam quis odio enim, in molestie libero. Vivamus cursus mi at \textit{nulla elementum sollicitudin}.

@@

Maecenas ultricies feugiat velit non mattis. Fusce tempus arcu id ligula varius dictum. 
\begin{itemize}
\item Curabitur pellentesque dignissim
\item Eu facilisis est tempus quis
\item Duis porta consequat lorem
\end{itemize}

@@

\begin{center}
\begin{tabular}{ccc}
%\includegraphics[width=0.4\linewidth]{harvard_univ_logo.png} & \hfill &
\includegraphics[width=0.2\linewidth]{seas-rgb.png}
\end{tabular}
\end{center}

>>>>

\usepackage{graphicx}  % Required for including images
\usepackage{hyperref}
\usepackage{natbib}

\usepackage{booktabs} % Top and bottom rules for tables
\usepackage{microtype}
\usepackage[linesnumbered,vlined,ruled]{algorithm2e} % algorithm

@@

We implement the parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
different solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for various complexities of ODEs from the basic exponential to
modeling sound wave propagation as we did in homework 4.

@@

\begin{block}{Algorithm}

The parareal algorithm, developed by Lions, Maday and Turinici in 2001, is
a general algorithm that allows parallelization in time.sIt does so by using a
cheaper (lower order, lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

Subscript: time index. Superscript: iteration index.

\begin{enumerate}
    \item Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$
    \item Compute the corrections $\delta g_n(u^1_n) =
        g_{\textnormal{fine}}(t_n, u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel
    \item Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta
        t}(t_n, u^2_n) + \delta g_n(u^1_n)$
    \item Repeat steps 2 and 3, incrementing the iteration label and using
        $u^{k+1}_0 = u^1_0$ as the initial condition
\end{enumerate}
\end{block}

@@

\begin{column}{\onecolwid} % Begin a column which is two columns wide (column 2)

%\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column

%\begin{column}{\onecolwid}\vspace{-.6in} % The first column within column 2 (column 2.1)

@@

@@

%\end{column} % End of column 2.1

%\begin{column}{\onecolwid}\vspace{-.6in} % The second column within column 2 (column 2.2)

@@

%\end{column} % End of column 2.2

%\end{columns} % End of the split of column 2 - any content after this will now take up 2 columns width

@@

%\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column again

%\begin{column}{\onecolwid} % The first column within column 2 (column 2.1)

@@

\begin{block}{Theoretical Results}

\textbf{Convergence}

Assume the coarse operator $g_{\Delta t}$ is order $m$ and is Lipshitz, and the
fine solution operator $g_{\textnormal{fine}}$ is a sufficiently accurate
approximation to the analytic operator so we may replace $g_{\textnormal{fine}}
\to g$.

\emph{Theorem}: The order of accuracy of the parareal method with coarse
solution operator $g_{\Delta t}$ and fine operator $g$ is $mk$. Can be proved by
induction.

\textbf{Stability}

With the parareal method, it is possible to combine ODE solvers. Stability
region depends on both $g_{\Delta t}$ and $g_{\textnormal{fine}}$, and the
equation being solved.

\textbf{Runtime and Speedup}

Let's say the coarse operator runs in time $t$, and the fine operator runs in
time $Q \cdot t$. Assume we have $N$ processors and we perform $k$ correction
steps. Then, the runtime of parareal is, assuming negligible setup and
aggregation time,

\begin{equation}
t + k(t + \frac{Qt}{N}).
\end{equation}

In order for there to be a speedup relative to the fine operator, we require
that $t + k(t + Qt/N) < Qt$, or $k < \frac{Q - 1}{1 + Q/N}$ or $N > \frac{Qk}{Q
- 1 - k}$.

We tested some stuff on Odyssey.

@@

\begin{column}{\sepwid}\end{column}

@@

\begin{block}{Preliminary Results}

\begin{figure}
\includegraphics[width=0.75\textwidth]{../data/runtime_vs_cpus.pdf}
\caption{Running time vs. number of CPUs.}
\label{fig:run_v_cpu}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{../data/error_vs_corrections.pdf}
\caption{Error vs. number of correction steps.}
\label{fig:err_v_k}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{../data/error_vs_qualityfactor.pdf}
\caption{Error vs. quality factor of the fine operator}
\label{fig:err_v_q}
\end{figure}

\begin{figure}
\includegraphics[width=0.75\linewidth]{../data/runtime_vs_corrections.pdf}
\caption{Running time as a function of the number of correction steps.}
\label{fig:run_v_k}
\end{figure}

@@

%\end{columns} % End of the split of column 2

%\end{column} % End of the second column

@@

@@

@@

\begin{center}
\includegraphics[width=0.2\linewidth]{seas-rgb.png}
\end{center}
